---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis-CI Build Status](https://travis-ci.org/pearcedom/srvALL.svg?branch=master)](https://travis-ci.org/pearcedom/srvALL)


# Installation
*survivALL* can be installed from R as follows

&nbsp;

```{r, eval = FALSE}
install.packages("survivALL")
```

&nbsp;

# Rationale

Survival analysis typically separates a patient cohort by some measure and compares the relative survival between groups. In practice, for a continuous measure, the decision of *where* to draw this division relies on previous knowledge or, more problematically, it may simply be a researchers best guess - for example separating around the median. Amongst other examples the associated vignettes outlined 3 alternative survival analysis approaches --- **median**, **hypothesis-driven** and our own **data-driven** approach as part of the `survivALL` package.


